### 2021-09-23 plos_latex_template.tex
some changes by Batool and Liz's revision and comments

Gathering a diverse team will only work if there is real inclusion. People with disabilities often say: `Nothing about us without us'; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups cannot be replaced by good intentions from people outside these groups. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. 
%liz, please check if this paragraph connects well with the whole rule. i am using nothing about us without us but also talking about all marginalized groups, in the context of building a team. att: andrea
% Rocío: Idea of this paragraph: real inclusion when having diverse members

@@

The following advice should apply to any kind of team, but are specially relevant when working with people from marginalized groups. First, rethink power relationships inside the group and share power. Be committed to examine power relations inside your organizing team and challenge it during all the process. % Rocío: Could we explain power relationship / share power ideas in more simple terms?
Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer leading positions, and let people from privileged groups step down. 
Build an environment in which every person has their voice and enforce voices from systematically excluded groups.
Offer mentorship and guidance if you encounter cases of impostor syndrome, doubts about the use of English in communication (See \textbf{Rule \ref{rule_language}}), or others. You can, for example, break the expectations about leadership as a lonely task and create smaller, co-led groups, where everyone finds their preferred tasks and gets to take leadership. Splitting the workload and responsibilities should not be done by putting care-taking labors--community building, meeting organization and note taking, conversations with potential partners--on the hands of women and other minoritized groups, while people from privileged groups take the lead in stereotypical highly-valued tasks (see \textbf{Rule \ref{rule_unbias}}). Make work visible, discuss collectively what should be done (and why), divide and account work. Being transparent and sharing the information is key to make work visible and collective. %sara: this relates to make labor visible from DF, maybe we can cite them here too. %Rocío: sure.
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work.

Most importantly, take care of your team. 
Having a diverse team and executing inclusive and accessible practices throughout the organizing period--it might be a year or two--may require a lot of effort. The effort is worth it because it strengthens the event and the community making it truly welcoming for everyone. 
However, having a strong community as the only reward may be enough for those who are in more privileged position. 
Some people and often the minoritized ones do not have institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free. 
In addition, tasks such as receiving and responding to code of conduct reports, can be emotionally intense work and should be additionally rewarded.
Prioritize your team's well-being. Check on them regularly and make sure everyone is comfortable. 
Be mindful of each particular context, be flexible with hours or commitments, and revise your budget (see \textbf{Rule \ref{rule_financial}}). 
% Rocío: Idea of this paragraph: take care of your team.

@@

Make sure that every selection committee--the committee looking for keynote speakers, the selection committee for abstracts, the prizes and award committees--are also diverse \cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}, and ask them to be aware that everyone has implicit biases, to recognize them, and try to counteract them. 
An inclusive and diverse organizing team (\textbf{Rule \ref{rule_organizing_team}}) is already a great starting point to overcome this bias in other roles. The regional and local communities in your field are also good sources to tap into. 
For example, for useR!2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential presenters, to co-lead social events, and attendees; some of them even held spaces to help the members of their groups to prepare abstracts for submission.
When organizing an international academic conference, you may reach local associations, student organizations, or other Early Career Researchers groups or communities of practice: ask for their requirements and barriers, and explicitly invite them to contribute to the conference. % Rocío: ask about their requirements and barriers sounds a bit confusing.
% Rocío: Not everyone knows what a community of practice is. Could we add a definition or use other words? Definition: "Communities of practice are groups of people who share a concern, a set of problems, or a passion about a topic, and who deepen their knowledge and expertise in this area by interacting on an ongoing basis". (Wenger, E., McDermott, R. A., & Snyder, W. (2002). Cultivating Communities of Practice: A Guide to Managing Knowledge. Harvard Business Press.)
Since many conferences are held with the primary goal of displaying cutting-edge work led by senior scientists and academics, early career researchers are often disadvantaged. Diversity should spread among all speaking roles, and the conference should also help raise the profiles of early career researchers, fostering collaboration and building their skills.

@@

@@

In-person interaction in conferences is priceless, but only for the ones who can afford to attend. 
Barriers such as cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications, are particularly true for global conferences that usually take place in high-income countries \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. 
% Rocío: we can add our Nature Correspondence letter to the citations later on. 
Online conferences are more inclusive: they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities \cite{salibaGettingGripsOnline2020}.
This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021, roosOnlineConferencesNew2020}.
% Rocío: we can add our Nature Correspondence letter to the citations here too. I would've liked one that talks about increase in participation of tutors and presenters but haven't found any. 
The online format may also make it easier to be inclusive of geographic regions by encompassing several timezones in the week or easily deciding to rotate the favored timezone year to year, without depending on a conference central location. 
% Rocío: main idea of the paragraph: online conferences are more inclusive

@@

Alternatively, a conference could have a hybrid format with an in-person and online component. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{ninerBetterWhomLeveling2021}.
Needless to say, an inclusive online component of a hybrid conference is not a free pass to have an exclusionary in-person component. % Rocío: I know this is a bit aggressive so I wouldn't mind if anybody helps with the wording. 
For instance, to avoid favoring the same regions in every edition, we suggest rotating the location of the conference—if it is global, moving from one continent to another. 
% Rocío: Hybrid conferences paragraph.

@@

% Rocío: Could start with an introduction (or put somewhere else) that online can easily make conferences more inclusive for many people, but not necessarily for people with disabilities, and that they can be easily left behind unless we actively work on their inclusion. 
Conferences are among the least accessible spaces that people with disabilities may encounter \cite{priceAccessImaginedConstruction2009}.
Ironically, accessibility practices are inclusive not only for people with disabilities but can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general.

If the conference is in person, the venue must be accessible for people with mobility limitations. 
Accommodations such as a quiet space for neurodivergent people should also be provided.
Presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. 
Regardless of the conference format --online, in person, or hybrid --images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. 
Platforms for conference registration and abstract submission, websites, and chat platforms--if used--should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). 
Captioning for presentations should be available in more than one language if possible.
Do not forget about making your social events accessible (e.g. art shows and movies) and making sure that the materials presented are accessible for all. 
These aspects should be tested well in advance of going live.

@@

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}. 
Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for wheelchairs in an in-person conference. 
On the other hand, it is likely that financial constraints may not allow you to make your conference completely accessible and there will be choices to make (e.g. choosing between captions and interpreters). 
Including people with disabilities, or rather, that are rendered disabled by standard practices in conferences in the organizing team would allow them to take part in the decisions from the beginning, spot right away the inaccessible practices that need to improve (see \textbf{Rule \ref{rule_organizing_team}}), and help make the right choices. As Sasha Costanza-Chock and the Design Justice League say, "center the voices of those who are directly impacted by the outcomes of the design process".  %Rocío: maybe slightly rephrase the last sentence since it seems to imply that everybody knows Sasha Costanza-Chock.

@@

Meeting the need of linguistically diverse audience is an important aspect in inclusion that is often overlooked. Non-native English speakers may feel intimidated to ask questions or raise their views. Providing a welcoming and diverse environment by encouraging the full participation of those individuals is critical (Rule 2). If the panelist or speakers will not be presenting in their native language, it may helpful to rephrase audience questions in a less complex language. Consider letting speakers know in advance that they can ask for a question to be translated to them in their native language and that they can respond in their native language.

@@

\begin{itemize}
    \item Promotion: Actively reach out and promote the conference to people who have been systematically excluded/Outreach/promotion plan that reaches communities that are traditionally excluded in scientific meetings
    \item Express this welcoming spirit in your communication strategy: Be public about your commitment to equity and inclusion: display the CoC prominently in conference spaces, publish the diversity statement, fee waivers, abstract selection process
    \item Languages
    \item Words matter: Be careful how you talk about minoritized people, Assertively set tone of inclusion
    \item coherence between actions and words
    \item \textbf{networking} when you gather a diverse groups of attendants you also have to take care of the spaces where they will interact
\end{itemize}

@@

%sara: this fits better in the communication rule
Set the tone. %adding here from Liz's rule but maybe this could go in communication in general
In addition to excluding derogatory or discriminatory language, make the effort to teach yourself the vocabulary and the best ways to communicate to account for every culture and situation. Do not expect people to teach you --it's not their role-- and accept feedback without being defensive.

\section{Allocate financial resources to support the presence of diverse people in every space of the conference
}
\label{rule_financial}

>>>>

Gathering a diverse team will only work if there is real inclusion. People with disabilities often say: `Nothing about us without us'; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups cannot be replaced by good intentions from people outside these groups. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. 
%liz, please check if this paragraph connects well with the whole rule. i am using nothing about us without us but also talking about all marginalized groups, in the context of building a team. att: andrea
% Liz: Yes!

% Rocío: Idea of this paragraph: real inclusion when having diverse members

@@

% Liz: I know this section is getting long, and maybe this is too blunt, but: 
% People who have experience with exclusion have the social and technical 
% expertise about removing barriers to inclusion.Actively ask where the barriers are and how to mitigate them.

The following advice should apply to any kind of team, but are specially relevant when working with people from marginalized groups. First, rethink power relationships inside the group and share power. Be committed to examine power relations inside your organizing team and challenge it during all the process. % Rocío: Could we explain power relationship / share power ideas in more simple terms?
Don't expect self-nomination and voting to work as mechanisms to counteract systemic inequalities. Nominate directly and offer leading positions, and let people from privileged groups step down. 
Build an environment in which every person has their voice and enforce voices from systematically excluded groups.
% Liz: Use of "voice" in this way makes deaf people feel excluded. Maybe it could be something like "Build an environment in which every person has input..." Or anything as long as it doesn't suggest you have to have a voice to participate.
Offer mentorship and guidance if you encounter cases of impostor syndrome, doubts about the use of English in communication (See \textbf{Rule \ref{rule_language}}), or others. You can, for example, break the expectations about leadership as a lonely task and create smaller, co-led groups, where everyone finds their preferred tasks and gets to take leadership. Splitting the workload and responsibilities should not be done by putting care-taking labors--community building, meeting organization and note taking, conversations with potential partners--on the hands of women and other minoritized groups, while people from privileged groups take the lead in stereotypical highly-valued tasks (see \textbf{Rule \ref{rule_unbias}}). Make work visible, discuss collectively what should be done (and why), divide and account work. Being transparent and sharing the information is key to make work visible and collective. %sara: this relates to make labor visible from DF, maybe we can cite them here too. %Rocío: sure.
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work.

% Liz: I have a little concern about mentoring WRT English, but I'm a white English speaker, so take it with a grain of salt. I think people who look like me need to do more accepting and less mentoring around English. If someone explicitly wants my help, I'm happy to give it because I love words and language, and I might learn something cool about another language. But I'm not sure we should be telling the English-priveleged readers that they are language mentors.

Most importantly, take care of your team. 
Having a diverse team and executing inclusive and accessible practices throughout the organizing period--it might be a year or two--may require a lot of effort. The effort is worth it because it strengthens the event and the community making it truly welcoming for everyone. 
However, having a strong community as the only reward may be enough for those who are in more privileged position. 
Some people and often the minoritized ones do not have institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free. 
% Liz: Yes! This is especially true when *some* people are paid to work on the conference. I was just asked to do accessibility work that should have been done by the contracted, paid organizer. This is hugely problematic and will be true for lots of conferences outside the open software community.
In addition, tasks such as receiving and responding to code of conduct reports, can be emotionally intense work and should be additionally rewarded.
Prioritize your team's well-being. Check on them regularly and make sure everyone is comfortable. 
Be mindful of each particular context, be flexible with hours or commitments, and revise your budget (see \textbf{Rule \ref{rule_financial}}). 
% Rocío: Idea of this paragraph: take care of your team.
% Liz: also, include people from diverse backgrounds in all aspects of planning, not just the diversity-related ones, unless that is their only interest. In some organizations, I could be on the scientific committee and participate in many other ways, but I'm only called on for accessibility.

@@

Make sure that every selection committee--the committee looking for keynote speakers, the selection committee for abstracts, the prizes and award committees--are also diverse \cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}, and ask them to be aware that everyone has implicit biases, to recognize them, and try to counteract them. 
An inclusive and diverse organizing team (\textbf{Rule \ref{rule_organizing_team}}) is already a great starting point to overcome this bias in other roles. The regional and local communities in your field are also good sources to tap into. 
For example, for useR!2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential presenters, to co-lead social events, and attendees; some of them even held spaces to help the members of their groups to prepare abstracts for submission.
When organizing an international academic conference, you may reach local associations, student organizations, or other Early Career Researchers groups or communities of practice: ask for their requirements and barriers, and explicitly invite them to contribute to the conference. % Rocío: ask about their requirements and barriers sounds a bit confusing.
% Rocío: Not everyone knows what a community of practice is. Could we add a definition or use other words? Definition: "Communities of practice are groups of people who share a concern, a set of problems, or a passion about a topic, and who deepen their knowledge and expertise in this area by interacting on an ongoing basis". (Wenger, E., McDermott, R. A., & Snyder, W. (2002). Cultivating Communities of Practice: A Guide to Managing Knowledge. Harvard Business Press.)
Since many conferences are held with the primary goal of displaying cutting-edge work led by senior scientists and academics, early career researchers are often disadvantaged. Diversity should spread among all speaking roles, and the conference should also help raise the profiles of early career researchers, fostering collaboration and building their skills.
% Liz: maybe consider a broader group than "early career researchers." One problem is that marginalized people have many fewer opportunities, and as they age that attenuation of opportunities compounds. Someone can be in the margins even years after finishing grad school or being a postdoc (not to mention that people without degrees could have contributions.

@@

% Liz: I might be able to, but I'm not sure where you want the additional text; where the beginning of the paragraph is because of how this is formatted. I couldn't agree more that being truly inclusive means not restricting pepole's role. Every subteam should be diverse, in addition to the whole organizing team. It's not just about representation but needs to encompass integration across the whole event.

@@

In-person interaction in conferences is priceless, but only for the ones who can afford to attend. 
Barriers such as cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications, are particularly true for global conferences that usually take place in high-income countries \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. 
% Rocío: we can add our Nature Correspondence letter to the citations later on. 
Online conferences are more inclusive: they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities \cite{salibaGettingGripsOnline2020}.
This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021, roosOnlineConferencesNew2020}.
% Rocío: we can add our Nature Correspondence letter to the citations here too. I would've liked one that talks about increase in participation of tutors and presenters but haven't found any. 
The online format may also make it easier to be inclusive of geographic regions by encompassing several timezones in the week or easily deciding to rotate the favored timezone year to year, without depending on a conference central location. 
% Rocío: main idea of the paragraph: online conferences are more inclusive
% Liz: but only if you make them that way. The R conference was a standout and innovative compared to others I've considred attending since Covid.

@@

Alternatively, a conference could have a hybrid format with an in-person and online component. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{ninerBetterWhomLeveling2021}.
Needless to say, an inclusive online component of a hybrid conference is not a free pass to have an exclusionary in-person component. % Rocío: I know this is a bit aggressive so I wouldn't mind if anybody helps with the wording. 
% Liz: I like it, but you could say
% an inclusive online component doesn't eliminate the need to organize an inclusive in-person component.

For instance, to avoid favoring the same regions in every edition, we suggest rotating the location of the conference—if it is global, moving from one continent to another. 
% Rocío: Hybrid conferences paragraph.

@@

% Rocío: Could start with an introduction (or put somewhere else) that online can easily make conferences more inclusive for many people, but not necessarily for people with disabilities, and that they can be easily left behind unless we actively work on their inclusion. 
Conferences are among the least accessible spaces that people with disabilities may encounter \cite{priceAccessImaginedConstruction2009}.
Ironically, accessibility practices are inclusive not only for people with disabilities but can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general. 
% Liz: I don't think it's actually ironic. maybe surprising or counterintuitive?

If the conference is in person, the venue must be accessible for people with mobility limitations. 
$ Liz: let's stay away from deficit-model language like "limitations." Possible aternatives:
% people who use wheelchairs or other assistive mobility devices
% ... must be physically accessible for disabled people

Accommodations such as a quiet space for neurodivergent people should also be provided.
% Liz: I hate "accommodateions because it sounds like something optional or for comfort or luxurious Consider something like "facilities"?

Presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. 
% Liz: avoid using disabilities as nouns like -- "the blind", "the deaf" etc. I woul duse "people who are deaf or hard-of-hearing" above. There's controversy about whether the word person
% should come first ("person who is blind") or the identity should come first ("blind person:) but the person should always be acknowledged (like "participants with disabilities", blind scientist", etc.. It's kind of objectifying without a person-related word.
Regardless of the conference format --online, in person, or hybrid --images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. 
% Liz: not just alt-text, the whole site should be screen-reader accessible so:
% communication about the conference should be fully screen-reader accessible including alt-text, and any videos should have both captions and sound.

Platforms for conference registration and abstract submission, websites, and chat platforms--if used--should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). 
Captioning for presentations should be available in more than one language if possible.
Do not forget about making your social events accessible (e.g. art shows and movies) and making sure that the materials presented are accessible for all. 
These aspects should be tested well in advance of going live.

@@

% Liz: It is essential to provide an accessible version of the slides for people who won't be seeing them One of the most important assumptions about presentations are that they will be both seen and heard. Usually, neither of these channels is a complete source of information about the presentation. For example, when discussing computer programming, code is usually displayed on slides while the presenter talks about them. They seldom read the details of what appears on the slides in enough detail for a blind person to learn, and a deaf person will also miss some explanation. Presenters  should provide as much information as possible on both channels, but in order to compensate for the fact that this is not common practice, blind people need to be able to read an accessible version of the slides at the same time they listen to the presentation. It is critical to make these materials available before the conference starts..

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}. 
Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for wheelchairs in an in-person conference. 
% Liz: "for people who use whellchairs" rather than "for wheelchairs"

On the other hand, it is likely that financial constraints may not allow you to make your conference completely accessible and there will be choices to make (e.g. choosing between captions and interpreters). 
Including people with disabilities, or rather, that are rendered disabled by standard practices in conferences in the organizing team would allow them to take part in the decisions from the beginning, spot right away the inaccessible practices that need to improve (see \textbf{Rule \ref{rule_organizing_team}}), and help make the right choices.
% Liz: maybe strengthen the last phrase to use their expertise to solve accessibility problems and collaborate on decisions about accessibility practices.

% Liz: As design justice author Sasha Costanza-Chock and the Design Justice League say, "center the voices of those who are directly impacted by the outcomes of the design process".  %Rocío: maybe slightly rephrase the last sentence since it seems to imply that everybody knows Sasha Costanza-Chock.
% not sure if that's a good characterization-- I am just learning about her right now.

@@

Meeting the need 
% Liz: People with disabilities don't want to be considered "needy". We reject the term "special needs" in part because it frames the person with the disability as being the burden causing the accommodation, where maybe it's the responsibility of society to be open to everyone. So I wonder if it's the same for people who speak other languages. Maybe they are not needy or inferior.  Or maybe it's wrong for me to apply disability rights philosophy to language.
% The linguistic diversity of conference participants is often overlooked,
% resulting in missed opportunities for them, as well as for the conference
% to benefit from their contributions.

of linguistically diverse audience is an important aspect in inclusion that is often overlooked.

 Non-native English speakers may feel intimidated to ask questions or raise their views. Providing a welcoming and diverse environment by encouraging the full participation of those individuals is critical (Rule 2). If the panelist or speakers will not be presenting in their native language, it may helpful to rephrase audience questions in a less complex language. Consider letting speakers know in advance that they can ask for a question to be translated to them in their native language and that they can respond in their native language.

@@

\begin{itemize}
    \item Promotion: Actively reach out and promote the conference to people who have been systematically excluded/Outreach/promotion plan that reaches communities that are traditionally excluded in scientific meetings
% Liz: including explicit descriptions of accessibility and other inclusion practices. Help people answer the question "Can I attend this conference?" without the emotional labor of contacting the organizers.
    \item Express this welcoming spirit in your communication strategy: Be public about your commitment to equity and inclusion: display the CoC prominently in conference spaces, publish the diversity statement, fee waivers, abstract selection process
    \item Languages
    \item Words matter: Be careful how you talk about minoritized people, Assertively set tone of inclusion
    \item coherence between actions and words
    \item \textbf{networking} when you gather a diverse groups of attendants you also have to take care of the spaces where they will interact
\end{itemize}

@@

%sara: this fits better in the communication rule
Set the tone. %adding here from Liz's rule but maybe this could go in communication in general
In addition to excluding derogatory or discriminatory language, make the effort to teach yourself the vocabulary and the best ways to communicate to account for every culture and situation. Do not expect people to teach you --it's not their role-- and accept feedback without being defensive.
$ Liz: should be "without being offended"
\section{Allocate financial resources to support the presence of diverse people in every space of the conference
}
\label{rule_financial}
